A Bioconductor workflow for processing and analysing spatial proteomics data

### 2018-04-03 bioc-workflow_march2017.Rnw
started to address comments

There are two aspects related to data normalisation that are relevant
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between channels without affecting
biological variability (i.e. the shape of the quantitative
profiles). This normalisation will depend on the underlying
quantitative technology and the experimental design, and will not be
addressed in this workflow. The second aspect, and more specific to
spatial proteomics data, is scaling all the organelle-specific
profiles into a same intensity interval (typically 0 and 1) by, for
example, dividing each intensity by the sum of the intensities for
that quantitative feature. This is not necessary in this example as
the intensities for each replicate have already been re-scaled to 1 in
Proteome Discoverer v1.4 Thermo Fisher. However, if the data require
normalisation, the user can execute the \texttt{normalise} function as
demonstrated in the below code chunk.

@@

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow
\cite{Webb-Robertson:2015,Lazar:2016}. In \cite{Gatto:2014b}, we have
described how a high content in missing values in spatial proteomics
data and their inappropriate handling leads to a reduction of
sub-cellular resolution. Missing data can be imputated using
\Biocpkg{MSnbase}'s \texttt{impute} function. The method underlying
the imputation method is then determined by a \texttt{methods}
parameter (see \texttt{?impute} for available options). To impute
missing values using nearest neighbour imputation, one would

@@

\begin{figure}[!ht]
  \centering
<<namap, out.width=".65\\textwidth">>=
image2(is.na(combined), col = c("black", "white"),
       main = "Missing values (white cells) after combining replicates")
@  
\caption{Heatmap of missing values. Note that the features are
  re-ordered to highlight cluster of proteins with similar numbers of
  missing values.}
  \label{fig:namap}
\end{figure}

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object. For example, if we wanted to highlight a proteins with the
accession numbers Q8CG48, Q8CG47, Q8K2Z4, and Q8C156, which are some
of the proteins known to form part of the 13S condensin complex, we
would create a first create a \texttt{FeaturesOfInterest} object, and
subsequently highlight their location on a PCA plot with the
\texttt{highlightOnPlot} function.

@@

\hl{Please note, in this document we can not evaluate the call to 
`phenoDisco` in the code chunk above as the algorithm is computationally 
intensive and best parallelised over multiple workers. This phenoDisco 
analysis took ~24 hours to complete when parallelised over 40 workers. 
The ouput of running the \texttt{phenoDisco} algorithm is an
\texttt{MSnSet} containing the new data clusters, appended to the
\texttt{featureData} under the name \texttt{pd}. We have made the results
readily available for users to interrogate and get an idea of the type
of output one may gain from using this function.} The results can be
displayed by using the \texttt{getMarkers} function. We see that 5 new
phenotype data clusters were found.

@@

% <<vishl, eval=FALSE>>=
% pRolocVis(hl, fcol = "pd")
% @

@@

<<setap, eval=FALSE>>=
par <- setAnnotationParams(inputs = c("Mus musculus", "UniProt/Swissprot"))

@

>>>>

There are two aspects related to data normalisation that are relevant
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between channels without affecting
biological variability (i.e. the shape of the quantitatives
profiles). This normalisation will depend on the underlying
quantitative technology and the experimental design, and will not be
addressed in this workflow. The second aspect, and more specific to
spatial proteomics data, is scaling all the organelle-specific
profiles into \sout{a}\hl{the} same intensity interval (typically 0 and 1) by, for
example, dividing each intensity by the sum of the intensities for
that quantitative feature. This is not necessary in this example as
the intensities for each replicate have already been re-scaled to 1 in
Proteome Discoverer v1.4 Thermo Fisher. However, if the data require
normalisation, the user can execute the \texttt{normalise} function as
demonstrated in the below code chunk.

@@

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow
\cite{Webb-Robertson:2015,Lazar:2016}. In \cite{Gatto:2014b}, we have
described how a high content in missing values in spatial proteomics
data and their inappropriate handling leads to a reduction of
sub-cellular resolution. \sout{Missing data can be imputated} \hl{We can impute missing data} using
\Biocpkg{MSnbase}'s \texttt{impute} function. The method underlying
the imputation method is then determined by a \texttt{methods}
parameter (see \texttt{?impute} for available options). To impute
missing values using nearest neighbour imputation, one would

@@

\begin{figure}[!ht]
  \centering
<<namap, out.width=".65\\textwidth">>=
image2(is.na(combined), col = c("black", "white"),
       main = "Missing values (white cells) after combining replicates")
@  
\caption{Heatmap of missing values. Note that the features are
  re-ordered to highlight \hl{clusters} of proteins with similar numbers of
  missing values.}
  \label{fig:namap}
\end{figure}

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object \hl{and then use the \textit{highlightOnPlot} function to 
visualise these}. For example, if we wanted to highlight \sout{a} 
proteins with the accession numbers Q8CG48, Q8CG47, Q8K2Z4, and Q8C156, 
which are some of the proteins known to form part of the 13S condensin complex, 
we \hl{would call the following code:}
\sout{would create a first create a \texttt{FeaturesOfInterest} object, and
subsequently highlight their location on a PCA plot with the
\texttt{highlightOnPlot} function.}

@@

\hl{Please note, in this document we can not evaluate the call to 
`phenoDisco` in the code chunk above as the algorithm is computationally 
intensive and best parallelised over multiple workers. This phenoDisco 
analysis took ~24 hours to complete when parallelised over 40 workers. 
The ouput of running the \texttt{phenoDisco} algorithm is an
\texttt{MSnSet} containing the new data clusters, appended to the
\texttt{featureData} under the name \texttt{pd}. We have made the results
readily available for users to interrogate and get an idea of the type
of output one may expect from using this function.} The results can be
displayed by using the \texttt{getMarkers} function. We see that 5 new
phenotype data clusters were found.

@@

<<vishl, eval=FALSE>>=
pRolocVis(hl, fcol = "pd")
@

@@

<<setap, eval=FALSE>>=
par <- setAnnotationParams(inputs = c("Mus musculus", "UniProt/Swissprot"))
@
